# Age Length Keys

1. Install R (http://cran.r-project.org/) packages and Tinn-R   (http://sourceforge.net/projects/tinn-r/).

```{r, eval = FALSE}
library(sp);library(mda);library(doBy);library(PBSmapping);library(maps);library(maptools);library(rgdal);library(spatial);library(fields)  # attach to R search path
```

2. Select some data from the ICES website - http://datras.ices.dk/Data_products/Download/Download_Data_public.aspx
3. Save the data to a directory of your choice, eg. 'c:/datras/data' and extract them from the zip files.
4. Install the R library datras. Do this by downloading datras*zip from the sharepoint. Open R, go to packages, install package from local zip files.
5. Attach datras package to search path
6. Extract the age length, sex, maturity data from the relevant directory

```{r}
#alk <- parseExchangeFormatALKs(wd = "data-raw")
alk <- all$CA %>% as.data.frame()
head(alk) # have a look at the data
```

7. Standardise to cm length categories
```{r}
alk$lngtclass[alk$lngtcode == "."] <- alk$lngtclass[alk$lngtcode == '.']/10
alk$lngtclass[alk$lngtcode == 0] <- alk$lngtclass[alk$lngtcode == 0]/10
```
8. Round down length classes & add 0.5
```{r}
alk$lngtclass[alk$lngtcode != "5"] <- round(alk$lngtclass[alk$lngtcode != "5"])
alk$lngtclass[alk$lngtcode != "5"] <- alk$lngtclass[alk$lngtcode != "5"]+0.5
print(table(alk$lngtclass)) # print distribution of ALL length categories
print(table(alk$age))      # print distribution of ALL ages
```
9. Put on species name (see Practical 2).  
```{r}
data(tsn)
alk$scientific.name <- as.character(tsn$completename[match(alk$speccode,tsn$tsn)])   # match scientific name onto species code 
sort(unique(alk$scientific.name)) # List of species with at least some age data
# Note: you tend to only get more commercially important species being aged etc.
table(alk$scientific.name,alk$age) # print distribution of ages by species (Note: this is for all ages combined).
```
10. Tidy up
```{r}
alk <- alk[!duplicated(alk),]     #  remove any duplicates
alk <- alk[alk$speccode != -9,]   #  missing species codes
alk <- alk[alk$lngtclass != -9,]  #  missing length classes
alk <- alk[alk$age != -9,]        # missing ages
alk <- alk[!is.na(alk$lngtcode),] #  missing length codes
alk <- alk[!is.na(alk$scientific.name),] # missing species codes
alk$age <- as.numeric(as.character(alk$age))
```
11. Select a species for a particular year and quarter and examine the age length key 
```{r}
table(alk$year,alk$scientific.name) # investigate range of years covered for each species
table(alk$quarter)                  # see range of quarters
```
12. Chose a species, year and quarter to examine
```{r}
alk$year <- as.integer(alk$year) # Needs to be fixed upstream
what.species    <- 'Pleuronectes platessa'   
what.year      <- 2010
what.quarter   <- 1

alk1 <- alk[alk$scientific.name == what.species & alk$year == what.year & alk$quarter == what.quarter,]

alk1t<-table(alk1$lngtclass,alk1$age) # create a table of the distribution of length by ages for the chosen species
 
print(alk1t)                          # have a look at the age distribution by length
 
rowSms <- apply(alk1t,1,sum,na.rm=T) # use apply to calculate row sums

alk1tp <- round(alk1t/rowSms,3)      # divide numbers by total to get proportions

print(alk1tp)                        # have a look at the proportions
```
13. Plot the data.
```{r, fig.width = 7.5, fig.height = 6.5}
# Try a barplot
par(mfrow=c(1,1))
barplot(t(alk1tp),space=0,col=1:dim(alk1tp)[[2]],ylab="P",xlab="Length(cms)",
legend.text=dimnames(alk1tp)[[2]],args.legend=list(x='topleft',bg='wheat'))

# Try a matrix plot 

matplot(alk1tp,type='l',ylab='P(age)',xlab="Length(cms)")

# Since data are a matrix why not try a 3D plot ?

alk1tp[alk1tp==0] <- NA
library(fields)
image.plot(as.numeric(dimnames(alk1tp)[[1]]), as.numeric(dimnames(alk1tp)[[2]]),alk1tp,col=topo.colors(100),xlab='Length(cms)',ylab='P(age)')
contour(as.numeric(dimnames(alk1tp)[[1]]), as.numeric(dimnames(alk1tp)[[2]]),alk1tp,add=T)
```
14. Apply ALK to Length frequency data

Simplest approach is to multiply the estimate of the proportion of each age at each length by the number of individuals at that length caught. If you do this you must ensure that you have a length in your ALK available for the corresponding length in your length frequency data.
```{r}
#lfs <- parseExchangeFormatLFs(wd="/home/einarhj/r/Pakkar/datrasr")
lfs <- all$HL %>% as.data.frame()
#  Again make sure to add on scientific name using TSN (downloaded from ITIS website). 
#data(tsn)  # itis database of all species in the world
#head(tsn)  # first 5 lines
lfs$scientific.name <- as.character(tsn$completename[match(lfs$speccode,tsn$tsn)])   # match scientific name 
# Tidy up
lfs <- lfs[!duplicated(lfs),]                  #  remove any duplicates
lfs <- lfs[lfs$speccode != -9,]                #  remove missing species codes
lfs <- lfs[lfs$lngtclass != -9,]               #  remove missing length classes
lfs <- lfs[!is.na(lfs$lngtclass),]             #  remove missing length classes
lfs <- lfs[!is.na(lfs$lngtcode),]              #  remove missing length codes
lfs <- lfs[!is.na(lfs$scientific.name),]       #  remove any missing scientific names 
lfs$hlnoatlngt <- lfs$hlnoatlngt*lfs$subfact   #  multiply by the subfactor
# Standardise length codes to cms (some are in mms).
lfs$lngtclass[lfs$lngtcode == "."] <- lfs$lngtclass[lfs$lngtcode == '.']/10
lfs$lngtclass[lfs$lngtcode == "0"] <- lfs$lngtclass[lfs$lngtcode == "0"]/10
lfs$lngtclass[lfs$lngtcode != "5"] <- round(lfs$lngtclass[lfs$lngtcode != "5"])  # Don't do this to fish binned at 5cm intervals
lfs$lngtclass[lfs$lngtcode != "5"] <- lfs$lngtclass[lfs$lngtcode != "5"]+0.5
# Useful to insert a weight for each length category for each species.
# Remember that this is usually done with parameters (a & b) from
# non-linear equations where Weight in grammes = aL^b  where L = length.
#data(length.weight)
# Match a and b parameters onto the correct species
data(length.weight)
lfs$a <- length.weight$a[match(lfs$scientific.name,length.weight$scientific.name)]
lfs$b <- length.weight$b[match(lfs$scientific.name,length.weight$scientific.name)]
# Use the parameters to estimate the weight in kgs.  Remember that weight = aL^b
lfs$hlwtatlngt<-((lfs$a*lfs$lngtclass^lfs$b)*lfs$hlnoatlngt) /1000
```

15. Extract a species to 'play' with. Remember that this should be one for which age length keys are available.

```{r}
plaice <- lfs[lfs$scientific.name == 'Pleuronectes platessa',]
```
16. Get the chrons (e.g. Practical01) because it's easier to use them to estimate effort at different spatial levels of aggregation.
```{r}
chrons <- all$HH %>% as.data.frame()
#  Usual 'tidy up' procedure.
chrons <- chrons[chrons$haulval=='V',]          # only valid hauls
chrons <- chrons[!duplicated(chrons),]          # chuck any duplicates
```
17. Make sure we have zeros in the dataset for stations that were trawled but no individuals caught.
```{r}
merged.plaice  <- mergeChronsLfs(chrons=chrons,length.frequencies=plaice)
fish <- merged.plaice 
```
18. Calculate a very simple index to see the principles using the year quarter combination selected above.
```{r}
lf1 <- fish[fish$year == what.year & fish$quarter == what.quarter,]  # select dat for what.year and what.quarter
head(lf1)
lf2 <- tapply(lf1$hlnoatlngt,lf1$lngtclass,sum,na.rm=T) # the total sum of individuals caught on that survey 
print(lf2) #Examine the data
# See if the available lengths correspond to the available aged lengths
length(lf2)
dim(alk1tp)
alk2tp <- data.frame(alk1tp)
dimnames(alk2tp)[[2]] <- c("lngtclass","age","p")   # Convert to a dataframe which can make life easier (relational database format)
alk2tp$lngtclass <- as.numeric(as.character(alk2tp$lngtclass))  # Make 'factors' into 'characters'
alk2tp$age <- as.numeric(as.character(alk2tp$age))              # 
head(alk2tp)
# Match the length frequencies with the lengths
alk2tp$hlnoatlngt <- lf2[match(alk2tp$lngtclass,as.numeric(dimnames(lf2)[[1]]))]
alk2tp$no.at.age  <- alk2tp$hlnoatlngt * alk2tp$p         # multiply the frequencies by the proportions
alk2tp$no.at.age[is.na(alk2tp$no.at.age)] <- 0            # replace Nas with zeros
# create the most straightforward index possible for that survey
index <- aggregate(list(no.at.age=alk2tp$no.at.age),list(age=alk2tp$age),sum,na.rm=T)
```

Plot the index for the year selected.
```{r, fig.width = 7.5, fig.height = 6.5}
par(mfrow=c(1,1))
plot(index$age,index$no.at.age,ylab='nos at age', xlab='age',type='b',lwd=1)  # Linear

plot(index$age,log(index$no.at.age+1),ylab='nos at age', xlab='age',type='b')   # Log scale
abline(lm(log(no.at.age+1)~age,data=index),lty=3,col='blue')
```
19. Now create a dataset of numbers by age per ICES statistical rectangle using the function NosatAgebyYearStatSquare
```{r, fig.width = 7.5, fig.height = 6.5}
out <- NosatAgebyYearStatSquare(lfdat = merged.plaice,alk = alk, chrons=chrons,  what.species ='Pleuronectes platessa') 
print(head(out))
# Plot the nos by year age and stat rectangle.
library(lattice)
xyplot(log(hlnoatage)~year|age,groups=statrec,type='l',data=out)
```
20. Compare indices calculated in different ways
```{r, fig.width = 7.5, fig.height = 6.5}
# total numbers divided by total effort
a <- aggregate(list(hlnoatage=out$hlnoatage), list(age=out$age,year=out$year),sum,na.rm=T) # sum the total numbers
hd <- aggregate(list(haulduration=chrons$hauldur/60), list(year=chrons$year),sum,na.rm=T)            # sum the effort 
b0 <- merge(a,hd,all=T)                        # merge
b0$cpue <- b0$hlnoatage/b0$haulduration        # calculate cpue
b0 <- b0[!is.na(b0$hlnoatage),]
print(head(b0)) # examine the data b0
# versus AVERAGE of the separate cpue in each statistical rectangle
out$cpue <- out$hlnoatage/out$haulduration
b1 <- aggregate(list(cpue=out$cpue), list(age=out$age,year=out$year),mean,na.rm=T)
print(head(b1))
# Plot them both to compare
par(mfrow=c(3,4),mar=c(2,2,3,2))     # This bit just finds the range of the data for plotting
u.ages <- sort(unique(b0$age))
r0 <- tapply(b0$cpue,b0$age,range)
r1 <- tapply(b1$cpue,b1$age,range)
r2 <- as.list(u.ages)
for (i in u.ages){
  r2[[i+1]] <- range(c(r0[as.character(i)],r1[as.character(i)])) 
}                                   
# Do the plotting
for(i in u.ages) {
  plot(b0$year[b0$age==i],b0$cpue[b0$age==i],type='l',xlab='',ylab='',ylim=r2[[i+1]])
  lines(b1$year[b1$age==i],b1$cpue[b1$age==i],type='l',col='blue')
  title(paste('age',i,sep=' '),cex.main=1.2)    }
```
20. Now create a dataset of numbers by age each station (eg. by lat and long) using the function NosatAgebyYearLonLat
```{r}
out <- NosatAgebyYearLonLat(lfdat = merged.plaice,alk = alk, chrons=chrons,  what.species ='Pleuronectes platessa') 
out[1:5,]                                        # First 5 lines
out$cpue.by.n <- out$hlnoatage/out$haulduration  # add on CPUE 
out$cpue.by.wt <- out$hlwtatage/out$haulduration
```
21. Plot data by age

On a bubble plots
```{r, fig.width = 7.5, fig.height = 6.5}
par(mfrow=c(1,1))
plotMapBlobs(out[out$age==3,],what.quarter=1,what.year = 2007,what.cpue='cpue.by.n',xlim0=c(-16,12),ylim0=c(45,62),scaling.factor=.5)
```
Or a grid
```{r, fig.width = 7.5, fig.height = 6.5, eval = FALSE}
grid.size.y <- .5                                                # grid size here is ICES statistical rectangle but this is entirely flexible
grid.size.x <- 1  

we <- -16                                                        # east and west boundaries
ea <- 10
so <- 48
no <- 62

par(mfrow=c(1,1))                                                # number of plots

out3 <- out[is.na(out$age) | out$age==3 & out$year ==2007,]


nos <-TrawlSurveyGridPlot(out3, we=we,ea=ea,so=so,no=no,nameLon = "shootlong", nameLat = "shootlat",plotMap=T,
nameVarToSum="hlnoatage",cellsizeX = grid.size.x, cellsizeY = grid.size.y,
legendx='bottomright',numcats=10,plotPoints=T,legendncol=2,paletteCats = "topo.colors",addICESgrid=TRUE,legendtitle="nos")
 
effort <-TrawlSurveyGridPlot(out3, we=we,ea=ea,so=so,no=no,nameLon = "shootlong", nameLat = "shootlat",plotMap=T,
nameVarToSum="haulduration",cellsizeX = grid.size.x, cellsizeY = grid.size.y,
legendx='bottomright',numcats=10,plotPoints=T,legendncol=2,paletteCats = "topo.colors",addICESgrid=TRUE,legendtitle="haul duration")
      
im.nos <-  as.image.SpatialGridDataFrame(nos, attr=2)      # extract nos from output of TrawlSurveyGridPlot
im.eff <-  as.image.SpatialGridDataFrame(effort, attr=2)   # extract effort from output of TrawlSurveyGridPlot

im.cpue <- im.nos
im.cpue$z <- im.nos$z/im.eff$z                             # divide catch by effort

#data(nseaBathy)                                            # North Sea
#data(nwestBathy)                                           # North west

library(fields)

image.plot(im.nos$x,im.cpue$y,log(im.cpue$z+1),col=topo.colors(10),xlab='',ylab='')                     # plot the data on a map
contour(nseaBathy$x,nseaBathy$y,nseaBathy$z,add=T,level=c(20,40,60),col='black')         # add on depth contour
#contour(nwestBathy$x,nwestBathy$y,nwestBathy$z,add=T,level=c(100,200,500,1000,2000,4000),col='black') 

maps::map("worldHires", add=TRUE, col='darkseagreen', fill=TRUE, bg="white",
regions=c('uk','ireland','france','germany','netherlands', 'norway','belgium',
'spain','luxembourg','denmark', 'sweden','iceland', 'portugal','italy','sicily','ussr','sardinia','albania','monaco','turkey','austria',
'switzerland','czechoslovakia','finland','libya', 'hungary','yugoslavia','poland','greece','romania','bulgaria', 'slovakia','morocco',
'tunisia','algeria','egypt' ))
```


Multi-nomial logit modeling of age length keys. The problem is that the data are packed full of missing lengths at various categories. The problem with the simple arithmetic approach above is that there are often no data available. There may be instances where data can be modeled over missing length categories to provide more information.
```{r}
alk <- all$CA %>% as.data.frame() #parseExchangeFormatALKs(wd="/home/einarhj/r/Pakkar/datrasr")
#Note: There is something wrong with the French IBTS data? 
#alk$lngtclas[alk$ship == 'THA2' & alk$year ==2006] <- alk$lngtclas[alk$ship == 'THA2' & alk$year ==2006]/10
alk$lngtclass[alk$lngtcode == "."] <- alk$lngtclass[alk$lngtcode == '.']/10
alk$lngtclass[alk$lngtcode == 0]   <- alk$lngtclass[alk$lngtcode == 0]/10
# Round it down
alk$lngtclass <- round(alk$lngtclass)
# Put on species name (see Practical 2).  
alk$scientific.name <- as.character(tsn$completename[match(alk$speccode,tsn$tsn)])   # match scientific name onto species code 
#  Tidy up
alk <- alk[!duplicated(alk),]     #  remove any duplicates
alk <- alk[alk$speccode != -9,]   #  missing species codes
alk <- alk[alk$lngtclass != -9,]  #  missing length classes
alk <- alk[!is.na(alk$lngtcode),] #  missing length codes
alk$age <- as.numeric(as.character(alk$age))
# Select a species for a particular year and quarter and examine the age length key 
# Chose a species, year and quarter to examine
what.species    <- 'Gadus morhua' # These are examples.
alk1 <- alk[alk$scientific.name == what.species,]
#Extract the columns of interest
alk2 <- data.frame(lngtclass=alk1$lngtclass,age=alk1$age,scientific.name = alk1$scientific.name)
xtabs(~age+lngtclass,data=alk2)     # examine the data
# install.packages('mlogit') # Install package if necessary
library(mlogit)
mldata <- mlogit.data(alk2,varying=NULL,choice='age',shape='wide')
mlogit.model<- mlogit(age~1|lngtclass, data = mldata, reflevel="1")
summary(mlogit.model)
#  Calculate the logit values
newdata <- data.frame(lngtclass = 1:90)
```

EINAR: TO DO - below is somewhat strange
```{r, eval = FALSE}
logit0 <- 13.9517 +  -0.8576*newdata$lngtclass
logit1 <- rep(0, 70)
logit2 <- -6.5674 +   0.1988*newdata$lngtclass
logit3 <- -10.099 +   0.2625*newdata$lngtclass
logit4 <- -15.8707 +  0.3558*newdata$lngtclass
logit5 <- -18.9503 +  0.3840*newdata$lngtclass
logit6 <- -19.8364 +  0.3791*newdata$lngtclass
logit7 <- -21.9500 +  0.3995*newdata$lngtclass
# We can exponentiate these logit values and scale them, knowing that the probabilities of for all the ages must sum to one.
logits <- cbind(logit0, logit1, logit2, logit3,logit4,logit5,logit6,logit7)
p.unscaled <- exp(logits)
p <- cbind(newdata, (p.unscaled / rowSums(p.unscaled)))
colnames(p) <- c("lngtclass", "age0", "age1", "age2", "age3","age4","age5","age6","age7")
apply(p[,-1],1,sum) #check

np <- p[,-1]

par(mfrow=c(1,1),mar=c(5,5,3,3))
barplot(t(np),space=0,col=1:dim(np)[[2]],ylab="P",xlab="Length(cms)",
        legend.text=dimnames(np)[[2]],args.legend=list(x='topleft',bg='wheat'))


# This basic approach can be extended to incorporate other 'predictor' variables, e.g. sex.

# As an exercise see if you can expand the model to do this and apply the predicted or modeled ALKs to the appropriate length-frequency data.


alk2 <- data.frame(lngtclass=alk1$lngtclass,age=alk1$age,sex=alk1$sex)

# Just males and females ano no missing data

alk3 <- alk2[alk2$sex %in% c('M','F'),]
alk3$sex <- ifelse(alk3$sex =="M",0,1)


xtabs(~age+lngtclass+sex,data=alk3)     # examine the data

mldata <- mlogit.data(alk3,varying=NULL,choice='age',shape='wide')

mlogit.model<- mlogit(age~1|lngtclass + sex, data = mldata, reflevel="1")

summary(mlogit.model)
```

